Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-02-02 analysis/paper/London_HauloutBeahvior.Rmd
change title
some edits for reviewer 1

title: Haul-out behavior and aerial survey detectability of seals in the Bering and Chukchi seas

@@

  In this paper, we use data collected from satellite-linked bio-loggers deployed
  between 2005 and 2020 to investigate the seasonal timing and environmental

@@

  ringed seals (_Phoca hispida_) are unique in their use of snow lairs, they were
  not included and a separate analysis is warranted. In addition to providing
  baseline data on phenology, these data also allow us to quantify 'availability',
  these seals basking on ice (i.e., to correct for the proportion of animals that
  are in the water while surveys are conducted). Using generalized linear mixed

@@

  to explain variation in hourly haul-out probability. We found evidence for

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have ripple effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering, Chukchi seas as a platform for important life history functions,
such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance of these species are unknown, so it is difficult to assess the effect,
if any, declines in sea ice habitat have had, or will have, on seal demography.

For ice-associated seals, haul-out behavior is tightly linked with sea ice.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a resting platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- bask on
sea-ice but also within snow lairs during winter and spring.

The remoteness of the Bering and Chukchi seas means direct scientific observation
of seal behavior is impractical. Thus, bio-logging devices are especially useful
tools for collecting key information on movement and haul-out behavior for these
species. Bio-logging records of time spent out of the water provide valuable
data for identifying covariates that explain variation in haul-out behavior. For
instance, in the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon. In the Arctic, Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer, and Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaked in May and June for bearded,
ribbon, and spotted seals.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
insufficient to permit strong inference about life history and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair would fall off) and, in
some situations, provide multiple years of data.

Ultimately, knowledge of trends in phenology and abundance (or life history
surrogates such as survival and recruitment) will be necessary to make credible
quantitative predictions about the effects of climate disruptions on the
abundance and distribution of Arctic seal populations. Before we can construct a
trend, however, we first require a baseline. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such abundance studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous issues
affecting the observation of seals on ice. These included availability (only
seals basking on ice were available to be counted), detection probability
(observers or automated detection systems may have missed some seals on ice),
species misclassification, and possible disturbance of seals by aircraft
[@conn2014a; @verhoef2014a]. Refining these inferences will improve the
accuracy of abundance estimates in the Arctic.

>>>>

title: Spring haul-out behavior of seals in the Bering and Chukchi seas

@@

  Improved estimates of seal population abundance are needed to properly monitor
  the impacts of these changes over time. In this paper, we use hourly
  percent-dry data from satellite-linked bio-loggers deployed
  between 2005 and 2020 to quantify the proportion of seals hauled out on ice,
  ice-associated seals (i.e., to correct for the proportion of animals that
  are in the water while surveys are conducted). In addition to providing
  essential data for survey 'availability' calculations, our analysis also provides 
  insights into the seasonal timing and environmental

@@

  ringed seals (_Phoca hispida_) can be out of the water but hidden from view in
  snow lairs, they were not included in this analysis. Using generalized linear mixed

@@

  to explain variation in haul-out probability. We found evidence for

@@

The global climate disruption is causing considerable reduction in Arctic sea
ice extent, volume, and seasonal presence [@kwok2018;
@meier2014; @overland2021; @wang2017]. These changes have tangible effects on
Arctic organisms, ecosystems, and the human communities who live in the region
[@huntington2020]. Such disruptions are a particular cause of concern for the
ice-associated seals that depend on spring and early summer sea ice (March-June)
in the Bering and Chukchi seas as a platform for important life history functions,
such as pupping, nursing, breeding behavior, and molting [@boveng2009;
@boveng2013a; @cameron2010; @kelly2010]. Limited data and large knowledge gaps
complicate predictions about the ultimate effects of changes in sea ice on the
behavior, health, abundance, and distribution of these seals. To date, indices
of seal health sampled during periods of declining sea ice differ regionally
[@crawford2015; @harwood2020]. Knowledge about evolutionary constraints on the
timing of reproductive and molting behavior is generally lacking, so it is
difficult to predict the readiness with which ice-associated seal species will
be able to adapt to future changes (e.g., by adjusting pupping or molting
schedules to earlier dates or different locales). This is further complicated by
the spatio-temporal variation in the phenology of these life history events
within regions and throughout their full ranges. Additionally, trends in
abundance of these species are unknown, so it is difficult to assess the effect,
if any, declines in sea ice habitat have had, or will have, on seal demography.

How ice-assoicated seals use sea ice as a haul-out platform varies between species.
Ribbon seals (_Histriophoca fasciata_) haul out of the water almost exclusively
on sea ice and are mostly pelagic outside the spring pupping, breeding, and
molting season [@boveng2018a]. While spotted (_Phoca largha_) and bearded
(_Erignathus barbatus_) seals rest on coastal features, they strongly prefer sea
ice as a haul-out platform during the spring and early summer [@frost2018].
Ringed seals (_Phoca hispida_) --- not included in this study --- haul out on
sea-ice but also within snow lairs during winter and spring.

The remoteness of the Bering and Chukchi seas means direct scientific observation
of seal behavior is impractical. Thus, bio-logging devices are especially useful
tools for collecting key information on movement and haul-out behavior for these
species. Bio-logging records of time spent out of the water provide valuable
data for identifying covariates that explain variation in haul-out behavior. 
For instance,Von Duyke et al.
[-@vonduyke2020] used satellite-linked bio-loggers to corroborate seasonal
changes between diurnal and nocturnal haul-out behavior of ringed seals
previously described by Kelly and Quakenbush [-@kelly1990c] using VHF radio tags
and direct observation. Bengtson et al. [-@bengtson2005a] documented a higher
propensity for ringed seal basking near solar noon, as did Ver Hoef et al.
[-@verhoef2014a] in an analysis of bearded, ribbon, and spotted seals using much
larger sample sizes. Olnes et al. [-@olnes2020] showed that the proportion of
time bearded seals spent hauled out progressively increased through spring and
summer, and Ver Hoef et al. [-@verhoef2014a] found haul-out probabilities
increased gradually starting in March and peaked in May and June for bearded,
ribbon, and spotted seals. Such analyses have not been limited to the Arctic. In 
the Antarctic, Bengtson and Cameron [-@bengtson2004] relied on
bio-logging data to demonstrate greater haul-out propensity in juvenile
crabeater seals (_Lobodon carcinophaga_) than adults, with highest probabilities
in February and at times close to solar noon.

Knowledge of haul-out patterns is not only important for understanding natural
history and ecology, but also for developing "availability" correction factors
for aerial surveys. Specifically, researchers need to know the fraction of seals
hauled out (versus in the water) when aerial surveys are conducted. Studies
estimating availability correction factors for seals typically use logistic
regression-style analyses to estimate the time-specific probability of being
hauled out based on 'wet/dry' data relayed by bio-loggers. In these models,
haul-out probabilities were expressed as a function of predictive covariates,
such as time-of-day, day-of-year, sex, age class, and environmental conditions
(e.g., [@reder2003], [@bengtson2004], [@bengtson2005a], [@udevitz2009],
[@verhoef2014a], and [@southwell2008]). However, sample sizes have often been
insufficient to permit strong inference about demographic and/or seasonal
variation in haul-out probabilities. For instance, Bengtson and Cameron's
[-@bengtson2004] study included 5 adult and 2 juvenile crabeater seals, while
Bengtson et al.'s [-@bengtson2005a] study was based on 6 ringed seals in the
Chukchi and Beaufort seas. These studies were often further limited by
logistical constraints on fieldwork and the attachment duration or operational
life of bio-loggers. In this study, we addressed some of these limitations by
deploying small bio-loggers designed for longer-term attachment on rear flippers
of a subset of the study individuals. These devices are designed to collect data
through the molt period (when those adhered to the hair would fall off) and, in
some situations, provide multiple years of data.

Statutory requirements under, for instance, the Marine Mammal Protection Act
(MMPA) and the Endangered Species Act (ESA) for timely estimates of population 
abundance and trend mean improved aerial survey effort must be paired with
improved knowledge of haul-out behavior. Several studies have contributed
estimates of the distribution and abundance of ice-associated seal species in
the Arctic using aerial surveys (e.g., [@bengtson2005a], [@conn2014a], and
[@verhoef2014a]). Such abundance studies were conducted over very large areas and
estimation of absolute abundance required making inference about numerous issues
affecting the observation of seals on ice. These included availability (only
seals on ice were available to be counted), detection probability
(observers or automated detection systems may have missed some seals on ice),
species misclassification, and possible disturbance of seals by aircraft
[@conn2014a; @verhoef2014a]. Refining these inferences will improve the
accuracy of abundance estimates and, hopefully, allow credible
predictions about the effects of climate disruptions on the abundance and
distribution of Arctic seal populations.
